---
title: "Introduction to Excerno"
author: "Audrey Mitchell, Marco Ruiz, Soua Yang"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{excerno-intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
--- 

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup}
# library(excerno)
```

# Introduction

Formalin-Fixation Paraffin-Embedding (FFPE) is a preservation technique for cancer tissue samples which introduces novel mutations. Leveraging the known mutational signature of FFPE and mutational signatures from the Catalogue of Somatic Mutations in Cancer (COSMIC) library, we set out to classify and filter FFPE artifacts. Our method utilizes non-negative matrix factorization (MutationalPatterns R package) and Bayes’ formula to calculate the probability that each mutation in a sample was caused by FFPE. Our methods are implemented in this package, excerno.

```excerno``` provide functions to help classify single nucleotide variants to their possible origin signature.

# Simulations

We tested the performance of our method by simulating mutations to match particular distributions and running a modified version of our classifier function that allows for the comparison of the true source of simulated mutations and the source predicted by our method. Functions for simulating mutations and classifying simulated mutations are included in the package.

## Creating simulated samples

Load the mutational signatures from COSMIC version 3 using MutationalPatterns.

```{r eval = FALSE}
library(MutationalPatterns)
cosmic.sigs <- get_known_signatures()
```

Extract COSMIC Signature 4 from the matrix of all COSMIC mutations. Create a vector of the 96 single base substitution mutation types using ````get_mutation_types``` and assign to the rownames of the Signature 4 matrix for compatibility with MutationalPatterns plotting functions. Use ```plot_96_profile``` from MutationalPatterns to visualize the distribution of Signature 4.

```{r eval = FALSE}
cosmic.sig4 <- as.matrix(cosmic.sigs[,4])
mutations <- get_mutation_types()
rownames(cosmic.sig4) <- mutations
plot_96_profile(cosmic.sig4)
```

Load the FFPE Signature using ```get_ffpe_signature```.

```{r eval = FALSE}
ffpe.sig <- get_ffpe_signature()
```

Create vectors of 100 mutations matching the distributions of COSMIC Signature 4 and the FFPE Signature using ```create_signature_sample_vector```.

```{r eval = FALSE}
sample.sig4 <- create_signature_sample_vector(cosmic.sig4, 100)
sample.ffpe <- create_signature_sample_vector(ffpe.sig, 100)
```

Use ```signature_cosine_similarity``` to calculate the cosine similarity between the simulated vectors above and the original signatures.

```{r eval = FALSE}
signature_cosine_similarity(sample.sig4, cosmic.sig4)
signature_cosine_similarity(sample.ffpe, ffpe.sig)
```

## Creating a classification data frame with simulated samples

Combine the Signature 4 and FFPE sample vectors and run the Bayesian classifier on the combined sample using ```classify_simulated_samples```.

```{r eval = FALSE}
# Turn into a list for input into classify_simulated_samples()
samples <- list(sample.sig4, sample.ffpe)
signatures <- list(cosmic.sig4, ffpe.sig)

classification.df <- classify_simulated_samples(samples, signatures)
classification.df
```

## Converting a simulated classification data frame to a GRange Object

The function ```create_gr_from_sample``` converts a classification data frame into a GRange Object. The classification data frame must have a mutations column and a truth column. The mutations column contains strings of mutations such as "A[C>T]A". The truth column indicates what mutational signature the mutation came from.

The output GRange object is meant to parallel a VCF file so additional columns are created in the GRange object: info, quality, filter, format, and samples (if provided values). These parameters are optional and will remain as NA if no values are provided. 

```{r eval = FALSE}
classification.gr <- create_gr_from_sample(classification.df, seq, "chr1")

# Adding values to other columns
info <- sample("SOMATIC", 200, replace = TRUE)
quality <- sample(50:100, 200, replace = TRUE)
filter <- sample("PASS", 200, replace = TRUE)
format <- sample("GT:GQ", 200, replace = TRUE)
samples <- list(sample(paste("0/0:", 1:100, sep = ""), 200, replace = TRUE), sample(paste("0/0:", 1:100, sep = ""), 200, replace = TRUE))
sample.names <- c("SAMPLE1", "SAMPLE2")

classification.gr <- data.frame(create_gr_from_sample(classify.df, seq, "chr1", info, quality, filter, format, samples, sample.names))
```

## Writing a classification data frame to a VCF file

The function ```write_grange_to_vcf``` takes in a GRange object and outputs a VCF file with the values from the GRange object.

```{r eval = FALSE}
vcf.filename <- "new_vcf.file"

write_grange_to_vcf(classification.gr, vcf.filename)
```

# Using excerno_vcf() on VCF files

## Inputs

### VCF file

```excerno_vcf``` takes VCF files as its main input source. Here is an example of loading in VCF files (included in package).

```{r eval = FALSE}
vcf.files <- list.files(system.file("extdata", package = "excerno"), pattern = "SIMULATED_SAMPLE_SBS4_\\d.vcf", full.names = TRUE)
```


### Artifact

To use ```excerno_vcf``` an artifact signature must be inputted.

### Method

```excerno_vcf``` offers two methods for calculating the contribution of signatures in a sample.

### Number of Signatures

### Target Signatures

```{r eval = FALSE}
excerno_vcf(vcf.files, artifact, "linear", target.sigs = target.sigs)
excerno_vcf(vcf.files, artifact, "nmf")
```








